---
title: "Constrained optimization"
author: "Pedro Amaral"
date: "2025-01-13"
output:
  html_document: default
  pdf_document: default
css: style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```


```{r, include=FALSE}
setwd("C:/Users/pamaral/CSU Fullerton Dropbox/Pedro Amaral/CSUF/Teaching/Econ 503/Resources/R")
```
One common definition of economics is that it is the study of how to best allocate scarce resources. Technically this is a process of constrained optimization (e.g., how to maximize utility subject to a limited income).

Consider the canonical example of a consumer that draws utility from the consumption of two goods $X$ and $Y$ according to a Cobb-Douglas utility function $U(X,Y)=X^{\alpha}Y^{1-\alpha},$ where $0<\alpha<1$. This consumer has an income of $m$ and the price of the two goods is $p_X$ and $p_Y,$ respectively. How much of each good should this consumer buy?

This problem can be expressed as:
$$ \max_{X,Y}\quad  U=X^{\alpha}Y^{1-\alpha}\\
  s.t. \quad p_X X\cdot p_Y Y \leq m.$$
  
There is more than way to solve this problem but we will concentrate on finding the optimal amounts by writing down the Lagrangian, taking first-order (necessary) conditions (FONC), and verifying the second-order (sufficient) conditions.

Write down the Lagrangian:

$\mathcal{L}(X,Y,\lambda)=X^{\alpha}Y^{1-\alpha}+\lambda\left(m-p_XX- p_YY \right)$

The FONC are:

$\frac{\partial \mathcal{L}}{\partial X} = \alpha \left(\frac{Y}{X}\right)^{1-\alpha}  -\lambda p_X =  0\\
     \frac{\partial \mathcal{L}}{\partial Y} = (1-\alpha) \left(\frac{X}{Y}\right)^{\alpha} -\lambda p_Y =  0\\
     \lambda \left(m-p_XX-p_XY\right) =  0\\
     \lambda \geq 0\\
     X p_X + Y p_Y \leq  m$

We can write the first two equations as 

$\lambda p_X = \alpha \left(\frac{Y}{X}\right)^{1-\alpha}\\
\lambda p_Y= (1-\alpha) \left(\frac{X}{Y}\right)^{\alpha}$

Dividing one by the other yields

$\frac{p_X}{p_Y}=\frac{\alpha}{1-\alpha}\frac{Y}{X},$

which we can solve for $Y$ as 

$Y=\frac{p_X}{p_Y}\frac{1-\alpha}{\alpha}X.$

Replacing this in the budget constraint with equality

$X p_X+ \left(\frac{p_X}{p_Y}\frac{1-\alpha}{\alpha}X\right) p_Y=m$

Simplifying and solving for $X$, yields

$X=\alpha \frac{m}{p_X},$

and solving for $Y$ using the equation above yields

$Y=(1-\alpha)\frac{m}{p_Y}.$

Suppose that $\alpha=0.5,$ $m=1,000$, $p_X=0.4,$ and $p_Y=0.6.$ Then the optimal solutions are

$X=\alpha \frac{m}{p_X}=0.5\frac{1,000}{0.4}=1,250\\
Y=(1-\alpha)\frac{m}{p_Y}=0.5\frac{1,000}{0.6}=833.(3)$


## Using R to solve constrained maximization problems
     
One of the routines we can use to solve a constrained optimization problem in R is `solnl` (but there are others). See [here](https://www.rdocumentation.org/packages/NlcOptim/versions/0.6/topics/solnl).

```{r}
# Call the library that contains the routine solnl
library(NlcOptim)

# define the problem's parameters 
alpha<-0.5
m<-1000
p_x<-0.4
p_y<-0.6

# initial guess for the optimal values
guess<-c(1,1)

# objective function
# specify the negative of the objective function if you want to maximize 
utility<-function(x){-x[1]^alpha*x[2]^(1-alpha)}

# now define the constraint
# start with a null space and use rbind to row bind them together
# then specify that we do not have an equality constraint ceq=NULL
# and we have an inequality constraint (<0)
constraint<-function(x){
  f=NULL
  f=rbind(f,p_x*x[1]+p_y*x[2]-m)
  return(list(ceq=NULL,c=f))
  }

optimal<-solnl(guess,utility,constraint)
```

Note that the solutions are exactly the same as the ones we obtained above analytically using the Lagrangian
```{r}
# print the results
sprintf("Optimal quantity of X is %.2f", optimal$par[1])
sprintf("Optimal quantity of Y is %.2f", optimal$par[2])


```



     
     