---
title: "Multivariate derivatives"
author: "Pedro Amaral"
date: "2025-01-10"
output:
  html_document: default
  pdf_document: default
css: style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```


```{r, include=FALSE}
setwd("C:/Users/pamaral/CSU Fullerton Dropbox/Pedro Amaral/CSUF/Teaching/Econ 503/Resources/R")
```
# Partial derivatives of multivariate functions

Consider the function $z=f(x,y).$

The partial derivative of $f$ with respect to $x$ at the point $(x_0,y_0)$ is $\frac{\partial f}{\partial x} (x_0,y_0)=\lim_{|\Delta x \rightarrow 0}\frac{f(x_0+\Delta x,y_0)-f(x_0,y_0)}{\Delta x}.$

The partial derivative $\frac{\partial f}{\partial y} (x_0,y_0)$is similarly defined.

We can think of $\frac{\partial f}{\partial x}(x,y)$ and $\frac{\partial f}{\partial y}(x,y)$ as functions of $x$ and $y.$ 

R can be used to take not only numerical derivatives, but also symbolic derivatives.

Take for example the function $z=f(x,y)=-x^2-2y^2+3x-2y+20.$

Let's plot it first.

```{r}
# Load libraries
library(ggplot2)
library(plotly)

# Generate data
x <- seq(-5, 5, length.out = 50)
y <- seq(-5, 5, length.out = 50)
f <- function(x, y) { -x^2-2*y^2 +3*x-2*y+20 }

# Create surface plot
plot_ly(x = x, y = y, z = outer(x, y, f), type = "surface")

```
## Numerical derivatives
Say we want to know the two first-derivatives at $(x_0,y_0)=(1,-1).$ Numerically, we can approximate it (albeit with some error) in the following way:
```{r}
# Define the point where you want to evaluate the derivative
x_0 <- 1
y_0 <- -1

# Create very small increments
delta <- 1e-6


# Numerically evaluate the partial derivative with respect to x
df_dx_num <- (f(x_0 + delta, y_0) - f(x_0, y_0)) / delta

# Print the result
print(df_dx_num)

# and do the same with respect to y
df_dy_num <- (f(x_0, y_0 + delta) - f(x_0, y_0)) / delta

# Print the result
print(df_dy_num)
```
## Symbolic derivatives

Alternatively, maybe you want to know what the partials $\frac{\partial f}{\partial x}$ and $\frac{\partial f}{\partial y}$ are.

R can deal with that too using the package "Deriv"
```{r}
library(Deriv)

# Calculate the partial derivative with respect to x
df_dx <- Deriv(f, "x") 
# print it
print(df_dx)

# Calculate the partial derivative with respect to y
df_dy <- Deriv(f,"y") 
# print it
print(df_dy) 

```
## Partial derivatives of a utility function

Consider the following utility function $u(c,h)=\frac{c^{1-\gamma}}{1-\gamma}-\psi\frac{h^{1+\theta}}{1+\theta}$ that depends on consumption $c$, and hours worked $h$, and has parameters $0<\gamma<1,$ $\psi>0,$ and $\theta>0.$ 

```{r}
u <- function(c,h,gamma,psi,theta) { c^(1-gamma)/(1-gamma)-psi*h^(1+theta)/(1+theta)}
```
We can calculate the partial derivatives, the marginal utilities, with respect to consumption and hours worked:
```{r}
# Calculate the partial derivative with respect to c
du_dc <- Deriv(u, "c") 
# print it
print(du_dc)

# Calculate the partial derivative with respect to h
du_dh <- Deriv(u, "h") 
# print it
print(du_dh)
```
Note how the marginal utility with respect to consumption is positive, while the marginal utility with respect to hours worked is negative.

We can now compute the second partial derivatives:
```{r}
# second derivative wrt consumption
d2u_dc2 <- Deriv(du_dc, "c") 
print(d2u_dc2)

# second derivative wrt hours worked
d2u_dh2 <- Deriv(du_dh, "h") 
print(d2u_dh2)

```
Note how the marginal utility of consumption is decreasing in consumption and the marginal utility of hours worked is decreasing in hours worked.

Finally we can compute the cross-derivatives:
```{r}
# cross-derivative of u_c wrt h
d2u_dcdh <- Deriv(du_dc, "h") 
print(d2u_dcdh)

# cross-derivative of u_h wrt c
d2u_dhdc <- Deriv(du_dh, "c") 
print(d2u_dhdc)
```
Note how the marginal utility of consumption is independent of hours worked and how the marginal utility of hours worked is independent of consumption.



