---
title: "Univariate derivatives"
author: "Pedro Amaral"
date: "2025-01-10"
output:
  html_document: default
  pdf_document: default
css: style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```


```{r, include=FALSE}
setwd("C:/Users/pamaral/CSU Fullerton Dropbox/Pedro Amaral/CSUF/Teaching/Econ 503/Resources/R")
```


# Derivatives of univariate functions

A derivative of a univariate function $y=f(x)$ at point $x_0$ is $f^\prime (x_0)=\lim_{\Delta \rightarrow 0}\frac{f(x_0+\Delta)-f(x_0)}{\Delta}.$

In words, it is the slope of the graph of $f$ at $(x_0,f(x_0).$

We can also think of $f^\prime (x)$ as a function of $x.$

R can be used to take not only numerical derivatives, but also symbolic (algebraic) derivatives (using libraries for that effect).

Take for example the quadratic function $f(x)=x^2+x-2.$

Let's plot it first
```{r}
# create a vector for x
x <- seq(-4, 3, length.out = 100)
# define the function
y <- function(x) x^2 + x - 2
# plot x versus f(x)
plot(x, y(x), type = "l", col = "blue", lwd = 2, xlab = "x", ylab = "y", main = "f(x) = x^2 + x - 2")

```

## Numerical derivatives

Say we want to know the derivative at $x_0=-2.$ Numerically, we can approximate it (albeit with some error) in the following way:
```{r}
# Define the point where you want to evaluate the derivative
x_0 <- -2

# Create a very small increment
delta_x <- 1e-6

# Numerically evaluate the derivative using the function y we defined above 
derivative <- (y(x_0 + delta_x) - y(x_0)) / delta_x

# Print the result
print(derivative)
```
## Symbolic derivatives

Alternatively, maybe you want to know what the derivative function  is.

R can help you here also. 

```{r}
# Install the Deriv package
#install.packages("Deriv")
library(Deriv)

# Calculate the derivative of y
d_y <- Deriv(y, "x") 

# print results
print(d_y) 
```
And we can also plot this derivative as a function of $x$ 
```{r}
plot(x, d_y(x), type = "l", col = "red", lwd = 2, xlab = "x", ylab = "y", main = "f'(x) = 2x+1")

```
Finally, we can confirm our numerical approximation to the derivative evaluated at $x_0=-2.$
```{r}
d_y(-2)
```
## Higher-order derivatives

With what we learned, it is trivial to obtain higher order derivatives. They are simply derivatives of derivatives, after all.
```{r}
# This time let's use a log function that we can keep on differentiating
z <- function(x) log(x)

# Calculate the first derivative of z
d_z <- Deriv(z, "x") 
print(d_z)

# Calculate the second derivative of z
d2_z <- Deriv(d_z, "x") 
print(d2_z)


# Calculate the third derivative of z
d3_z <- Deriv(d2_z, "x") 
print(d3_z)

```



