---
title: "Using FRED data in R"
output:
  html_document: default
css: ../style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```
## Pedro Amaral

## 2026-03-21

# Using FRED data in R

This is a simple example of how to use an API to import data in R. In this case we are using FRED's API and a package designed to import from it called `fredr`. You can get [practical details about this package here](http://sboysel.github.io/fredr/articles/fredr.html), as well as a [more complete description of the FRED API here](https://fred.stlouisfed.org/docs/api/fred/).  

After installing `fredr` start by calling it along with `tidyverse`.

```{r}
library(fredr)
library(tidyverse)
```

Note that you need to register with FRED (which is free) to get an API key to be able to do this.

Once you have an API key, you set it for the current session. You can also set it once and for all by modifying an environment file -- see the link above. 

Now you're ready to start importing from FRED. In this case we are downloading inflation and unemployment to plot a Phillips curve.

Importing CPI inflation

```{r}

# Set your FRED key
#fredr_set_key("################################")
fredr_set_key(Sys.getenv("FRED_API_KEY"))

cpi_df_raw<-fredr(
  series_id = "CPIAUCSL",
  observation_start = as.Date("1948-01-01"),
  observation_end = as.Date("2025-02-01"),
  frequency = "m", # monthly
  units = "pc1" # percentage change from a year ago
)
```

Importing the unemployment rate

```{r}
ur_df_raw<-fredr(
  series_id = "UNRATE",
  observation_start = as.Date("1948-01-01"),
  observation_end = as.Date("2026-02-01"),
  frequency = "m", # monthly
  units = "lin" # levels (percent)
)
```

This imports a tibble (a type of data frame). Here's what the data looks like:
```{r}
head(cpi_df_raw)
```
```{r}
head(ur_df_raw)
```


There are 5 variables but we are only interested in 2 of them: `date` and `value`, so let's save a subset of this data frame

```{r}
cpi_df<-subset(cpi_df_raw, select = -c(series_id,realtime_start,realtime_end) )
ur_df<-subset(ur_df_raw, select = -c(series_id,realtime_start,realtime_end) )
```

Let's rename inflation and unemployment because they are both called `value`

```{r}
names(cpi_df)[names(cpi_df)=="value"] <- "cpi"
names(ur_df)[names(ur_df)=="value"] <- "ur"
```

Now let's merge the two date frames into one to tidy things up. Note that we are merging by `date`.

```{r}
phillips_df <- merge(cpi_df,ur_df,by="date")
```

Here's what the new data frame looks like:

```{r}
head(phillips_df)
```

Finally, let's plot the Phillips curve. Note how the variables in a data frame are named

```{r, fig.alt="Scatterplot with the unemployment rate on the x-axis and the CPI inflation rate on the x-axis for the US, monthly, from 1948 to 2026."}
ggplot(phillips_df, aes(x=ur, y=cpi)) + 
  geom_point()+
  labs(title="Phillips curve", x="Unemployment rate (%)", y="CPI inflation (%)")
```



