---
title: "Using the Hodrick-Prescott filter in R"
output:
  html_document: default
css: ../style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```
## Pedro Amaral

## 2026-03-21

# The Hodrick-Prescott (HP) filter in R

This is a simple example of how to use the Hodrick-Prescott (HP) filter in R.

The HP filter is used to extract a trend component and a cyclical component from of a time series of raw data. 

The filter extracts a trend that minimizes the contemporaneous squared difference between data and trend plus the weighted squared differences in trend changes in consecutive periods. The weight $\lambda$ is specified by the user and reflects how averse the user is to short-run movements in the trend. [See more details here](https://en.wikipedia.org/wiki/Hodrick%E2%80%93Prescott_filter).

In R, the HP filter is part of a larger package of time-series filters: [the `mFilter` package](https://cran.r-project.org/web/packages/mFilter/index.html).

You can learn more about implementing the `hpfilter` function [in this explanation at CRAN](https://search.r-project.org/CRAN/refmans/mFilter/html/hpfilter.html).


```{r}
#install.packages("mFilter")
# Load library
library(mFilter)
```


Start by downloading real GDP from FRED so we can use it in this example.

```{r}
library(fredr)
library(tidyverse)
# Set your FRED key
#fredr_set_key("################################")
fredr_set_key(Sys.getenv("FRED_API_KEY"))

gdp_raw<-fredr(
  series_id = "GDPC1",
  observation_start = as.Date("1947-01-01"),
  observation_end = as.Date("2025-10-01"),
  frequency = "q", # quarterly
  units = "lin" # levels
)

# save the gdp values and the quarters 
gdp=gdp_raw$value
quarters=gdp_raw$date
```

Run the HP filter with the default value for a quarterly series of 1600 for lambda.

```{r}
hp <- hpfilter(gdp,freq=1600,type = "lambda")
```

This results in a data frame with the actual va;ue, the trend and a cyclcal component.

Save the trend and define the cycle as the percentage difference between the original series and the trend.

```{r}
trend<-hp$trend
cycle<-((gdp-trend)/trend)*100

# save everything in a data frame
df<-tibble(quarters,gdp,trend,cycle)
```

Plot everything.

```{r, fig.alt="The first figure shows quaterly GDP and H-P filtered trend from 1947 to 2025. The second figure shows the cyclical component of GDP."}

# Trend vs original series
ggplot(df, aes(x=quarters)) + 
  geom_line(aes(y = gdp)) + 
  geom_line(aes(y = trend), color="red", linetype="twodash")+
  labs(x="Quarters",y="Billions of Chained 2017 Dollars")


# Cyclical component
ggplot(df, aes(x=quarters)) + 
  geom_line(aes(y = cycle))+
  labs(x="Quarters",y="GDP (% deviation from trend)")

```




