---
title: "Cross-country income dispersion"
output:
  html_document: default
css: ../style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```

## Pedro Amaral

## 2026-01-24

## Cross-country income dispersion

Let's compute cross-country income dispersion measures using the Penn World Table (PWT) database.

Start by installing the package ``pwt10` that contains the Penn World Table 10 data base. The PWT is a database with information on relative levels of income, output, input, and productivity, covering 183 countries between 1950 and 2019. You can see more details at the [Groningen's center's website](https://www.rug.nl/ggdc/productivity/pwt/).


```{r}
#install.packages("pwt10")

# call the library
library(pwt10)

# get the data base that's called pwt10.0
data("pwt10.0")
```

From this data base lets construct a smaller one, with the following variables from 1950 to 2019:

1. `country`: country names `
2. `years`: years from 1950 to 2019 for each country
3. `rgdpe`: expenditure-side real GDP at chained PPPs (in mil. 2017US$):  
4. `pop`: population (in millions)

```{r}
# choose appropriate variables
df1 <- pwt10.0[,c(2,3,5,7)]
```

Now construct GDP per capita and keep only the names, years, and GDP per capita

```{r}
df1$gdp_pc<-df1$rgdpe/df1$pop
df1 <- df1[c(1,2,5)] 
```

Next, keep only 1970 and 2019

```{r}
# compute the number of years we have
n_years=max(df1$year)-min(df1$year)+1

# use the number of years and the number of total observations to compute the number of countries
n_countries=nrow(df1)/n_years

# use the dplyr library (to use the function filter)
library(dplyr)

# filter data only for 1970 and 2019
df1<- df1 %>% filter(year==1970|year==2019)
```

Create a new data frame with 3 columns, the country name, GDP per capita in 1970, and GDP per capita in 2019, with each observation being a different country and filter out countries for which we are missing observations, or are outliers

```{r}

# create the data frame
df2 <- data.frame(column1 = factor(n_countries), column2 = numeric(n_countries), column3 = numeric(n_countries))
colnames(df2) <- c(c("isocode","GDP_pc_1970","GDP_pc_2019")) 

# populate the name column
df2$isocode<-unique(df1[,1])

# populate the two GDP per capita columns 
for (i in 1:n_countries) {
  df2$GDP_pc_1970[i] = df1$gdp_pc[2*i-1]
  df2$GDP_pc_2019[i] = df1$gdp_pc[2*i]
}

# filter out NAs 
df2 <- df2 %>% filter(!is.na(df2$GDP_pc_1970)) 

# exclude outliers that report over 90,000 GDP per capita in 1970
df2 <- df2 %>% filter(df2$GDP_pc_1970<90000) 
```

Next, chart an histogram with the distribution of GDP per capita in 2019

```{r, fig.alt ="A histogram of cross-country GDP per capita in 2019"}
#install.packages("ggplot2")
library(ggplot2)

# histogram
ggplot(df2, aes(x=GDP_pc_2019)) + 
  geom_histogram(aes(y = ..density..),color="black", fill="white", binwidth = 5000)+
  geom_density(lwd = 1, color = "blue",
               fill = "blue", alpha = 0.25)+
  labs(x="GDP per capita 2019")+
  labs(alt="A histogram of cross-country GDP per capita in 2019")
```

We can also compute some measures of dispersion, like decile ratios (90/10 and 80/20)


```{r}
# compute deciles cutoff
deciles_2019 <- quantile(df2$GDP_pc_2019, c(0.1, 0.2, 0.8, 0.9))
deciles_1970 <- quantile(df2$GDP_pc_1970, c(0.1, 0.2, 0.8, 0.9))


# compute means within groups
bot_10_2019 <- mean(df2$GDP_pc_2019[df2$GDP_pc_2019 < deciles_2019[1]])
bot_20_2019 <- mean(df2$GDP_pc_2019[df2$GDP_pc_2019 < deciles_2019[2]])

bot_10_1970 <- mean(df2$GDP_pc_1970[df2$GDP_pc_1970 < deciles_1970[1]])
bot_20_1970 <- mean(df2$GDP_pc_1970[df2$GDP_pc_1970 < deciles_1970[2]])

top_80_2019 <- mean(df2$GDP_pc_2019[df2$GDP_pc_2019 > deciles_2019[3]])
top_90_2019 <- mean(df2$GDP_pc_2019[df2$GDP_pc_2019 > deciles_2019[4]])

top_80_1970 <- mean(df2$GDP_pc_1970[df2$GDP_pc_1970 > deciles_1970[3]])
top_90_1970 <- mean(df2$GDP_pc_1970[df2$GDP_pc_1970 > deciles_1970[4]])


# compute ratios
t9_2019=top_90_2019/bot_10_2019
t9_1970=top_90_1970/bot_10_1970


t8_2019=top_80_2019/bot_20_2019
t8_1970=top_80_1970/bot_20_1970


sprintf("The 90 to 10 ratio has grown from %.2f in 1970 to %.2f in 2019.", t9_1970, t9_2019)

sprintf("The 80 to 20 ratio has grown from %.2f in 1970 to %.2f in 2019.", t8_1970, t8_2019)


```

Let's see which countries did particularly well when it comes to growth (miracles) and particularly poorly (disasters)


```{r, fig.alt ="A scatterplot of cross-country GDP per capita in 1970 vs 2019"}
# identify miracles and disasters
i_miracles=which(df2$GDP_pc_2019/df2$GDP_pc_1970>=9)
i_disasters=which(df2$GDP_pc_2019/df2$GDP_pc_1970<1)

# lines for miracles and disasters
diag_x <- c(0,max(df2$GDP_pc_1970))
diag_disaster <- diag_x
diag_miracle <- 9*diag_x


# chart
ggplot(df2, aes(x=GDP_pc_1970, y=GDP_pc_2019)) + 
  geom_point()+
  geom_abline(intercept = 0, slope = 9, color = "blue", lwd=1,linetype = "dashed")+
  geom_abline(intercept = 0, slope = 1, color = "red",lwd=1,linetype = "dashed")+
  geom_text(x=df2$GDP_pc_1970[149]+1000, y=df2$GDP_pc_2019[149], label=df2$isocode[149],size=3)+
  geom_text(x=df2$GDP_pc_1970[108], y=df2$GDP_pc_2019[108]-2000, label=df2$isocode[108],size=3)+
  geom_text(x=df2$GDP_pc_1970[69], y=df2$GDP_pc_2019[69]+3000, label=df2$isocode[69],size=3)+
  geom_text(x=df2$GDP_pc_1970[81]+1000, y=df2$GDP_pc_2019[81], label=df2$isocode[81],size=3)+
  geom_text(x=df2$GDP_pc_1970[23]-1000, y=df2$GDP_pc_2019[23], label=df2$isocode[23],size=3)+
  labs(x="GDP per capita in 1970") +
    labs(y="GDP per capita in 2019")+
  labs(alt="A scatterplot of cross-country GDP per capita in 1970 vs 2019")
 
```

