---
title: "The simple Solow model"
output:
  html_document: default
css: ../style.css
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```

## Pedro Amaral

## 2026-01-24

## The simple Solow model

In this first approach to the Solow model we start by assuming that population and and total factor productivity (TFP) are constant over time. Consider a Cobb-Douglas production function $y_t=A k_t^{\alpha}$ in capital per worker.

```{r}
# Set parameters

# Total factor productivity
A=10

# Capital's share of income
alpha=0.4

# Depreciation
delta=0.1

# Saving rate
s=0.25
```

Let's fist solve for steady-state numerically by finding the zero for the capital accumulation equation $k_{t+1}-k_t=s A k_t^{\alpha}-\delta k_t=0.$ To do this let's use the function `uniroot` that finds the zero of a univariate function in a given interval. Recall this function has two zeros, one of them being $k=0$, which is the trivial steady-state. To make sure we don't get that back, we can tell `uniroot` to search for a zero within an interval that excludes zero. Just make sure the endpoints of your interval have different signs to make sure the function crosses the y-axis.

```{r}
# Define the function
f<-function(k) {s*A*k^alpha-delta*k}

# Call the root finder
zero_k<-uniroot( f, c(1,300) )
k_star<-zero_k$root
```

Next, let's obtain a path for capital per worker assuming we start somewhere below steady state

```{r}
# Initial period capital 
k=10

# Define number of periods you want the economy to run for
T=99

# Loop for each period
for(i in 1:T){
  k[i+1] <- (s*A*k[i]^alpha+(1-delta)*k[i])
}
```

Let's plot this capital path

```{r, fig.alt="A plot of the evolution of capital per worker over time"}
# chart
library(ggplot2)

# chart
library(ggplot2)

# create a data fram just to make thngs easier
periods <- 1:(T+1)
df<-data.frame(periods,k)

p1<-ggplot(df,aes(x = periods, y=k)) +
  geom_line()+
  geom_hline(yintercept = k_star, color = "red", linetype = "dashed")+
  labs(x = "Periods", y="Capital per worker", title = "Capital")

p1
```

With the capital per worker path in hand, we can obtain paths for all other variables in the model

```{r}
# output per worker
y=A*k^alpha
y_star=A*k_star^alpha

# consumption per worker
c=(1-s)*y
c_star=(1-s)*y_star

# investment per worker
i=s*y
i_star=s*y_star

# rental rate of capital
R=A*alpha*k^(alpha-1)
R_star=A*alpha*k_star^(alpha-1)

# wages
w=y-k*R
w_star=y_star-k_star*R_star

# add all this to data frame
df$y <- y
df$c <- c
df$i <- i
df$R <- R
df$w <- w
```

We can now plot all the variables together to appreciate the whole economy converging to steady-state

```{r, fig.alt="Six sub-plots with the evolution of model variables over time"}

p2<-ggplot(df,aes(x = periods, y=y)) +
  geom_line()+
  geom_hline(yintercept = y_star, color = "red", linetype = "dashed")+
  labs(x = "Periods", y="Output per worker", title = "Output")

p3<-ggplot(df,aes(x = periods, y=c)) +
  geom_line()+
  geom_hline(yintercept = c_star, color = "red", linetype = "dashed")+
  labs(x = "Periods", y="Consumption per worker", title = "Consumption")

p4<-ggplot(df,aes(x = periods, y=i)) +
  geom_line()+
  geom_hline(yintercept = i_star, color = "red", linetype = "dashed")+
  labs(x = "Periods", y="Investment per worker", title = "Investment")

p5<-ggplot(df,aes(x = periods, y=R)) +
  geom_line()+
  geom_hline(yintercept = R_star, color = "red", linetype = "dashed")+
  labs(x = "Periods", y="Rental rate of capital", title = "Rental rate of capital")

p6<-ggplot(df,aes(x = periods, y=w)) +
  geom_line()+
  geom_hline(yintercept = w_star, color = "red", linetype = "dashed")+
  labs(x = "Periods", y="Wages", title = "Wages")

# Arrange the plots in a grid
library(gridExtra)

p<- grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)

```
