---
title: "ECON 503 Textbook Exercises — Chapter 27"
author: "Pedro Amaral"
date: today
format:
  html:
    toc: true
    code-fold: true
    css: style_C.css
    embed-resources: true
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tibble)
library(gridExtra)

# ── Shared panel theme ────────────────────────────────────────────────────────
th_tp <- theme_minimal(base_size = 11, base_family = "serif") +
  theme(
    plot.background  = element_rect(fill = "#fdfcf8", color = NA),
    panel.background = element_rect(fill = "#fdfcf8", color = NA),
    panel.grid       = element_blank(),
    axis.line.x      = element_line(color = "black", linewidth = 0.5),
    axis.line.y      = element_line(color = "black", linewidth = 0.5),
    axis.ticks.x     = element_line(color = "black", linewidth = 0.4),
    axis.ticks.y     = element_blank(),
    axis.text.y      = element_blank(),
    axis.text.x      = element_text(size = 10, color = "black", family = "serif"),
    axis.title.x     = element_text(hjust = 1, size = 11, family = "serif"),
    axis.title.y     = element_text(hjust = 1, angle = 0, vjust = 1,
                                    size = 11, family = "serif"),
    plot.margin      = margin(10, 14, 10, 38)
  )

# ── Helpers ───────────────────────────────────────────────────────────────────

# Build the data frame of horizontal segments for a step-function time series.
# vals   : level in each epoch (length n)
# breaks : times at which the level changes (length n-1)
# t_max  : right edge of the plot
make_hsegs <- function(vals, breaks, t_max = 5.4) {
  t_L <- c(0, breaks)
  t_R <- c(breaks, t_max)
  tibble(x = t_L, xend = t_R, y = vals, yend = vals)
}

# Draw one time-series panel.
# segs        : output of make_hsegs — drawn as solid lines
# ylab        : y-axis title (expression or string)
# ref_labels  : list of lists, each with $y (numeric) and $lab (expression/string)
#               — placed as text just left of the y-axis
# vert_lines  : x-positions of dotted vertical lines
# xlim, ylim  : axis limits; ylim must leave room below 0 for the labels
plot_tp <- function(segs, ylab, ref_labels,
                    vert_lines = c(1, 3),
                    xlim       = c(0, 5.4),
                    ylim       = c(-18, 105),
                    lw         = 1.3) {

  p <- ggplot()

  # dotted vertical lines at event times
  for (xv in vert_lines)
    p <- p + geom_vline(xintercept = xv, linetype = "dotted",
                        color = "gray40", linewidth = 0.55)

  # solid horizontal segments (step levels)
  p <- p + geom_segment(data = segs,
                        aes(x = x, xend = xend, y = y, yend = yend),
                        linewidth = lw, color = "black", lineend = "square")

  # left-side level labels
  for (item in ref_labels)
    p <- p + annotate("text",
                      x = xlim[1] - 0.16, y = item$y,
                      label = item$lab,
                      size = 3.25, hjust = 1, vjust = 0.5, family = "serif")

  p +
    scale_x_continuous(breaks = c(1, 2, 3), labels = c("1", "2", "3"),
                       limits = xlim, expand = c(0, 0)) +
    scale_y_continuous(limits = ylim, breaks = NULL, expand = c(0, 0)) +
    labs(x = "Time", y = ylab) +
    coord_cartesian(clip = "off") +
    th_tp
}

# ── Calibration (normalised [0, 100] space) ───────────────────────────────────
theta0 <- 28;  theta1 <- 72          # permanent θ increase

# Pre-shock baseline (original NCL equilibrium)
Y0 <- 70;  P0 <- 32;  r0 <- 25

# Long-run NCL equilibrium after permanent θ increase (same for both models)
Y_ncl <- 30;  P_ncl <- 75;  r_ncl <- 68

# PSP halfway values (midpoint between original and NCL)
Y_mr <- (Y0 + Y_ncl) / 2   # 50
P_mr <- (P0 + P_ncl) / 2   # 53.5
r_mr <- (r0 + r_ncl) / 2   # 46.5
```

---

# Exercise 27.1 — Increase in $\theta_t$

**Problem.** Suppose the economy starts in the Neoclassical equilibrium and
$\theta_t$ increases.

**(a)** Draw the dynamics in the simple sticky-price case. Verbally describe
what is going on.

**(b)** Draw the dynamics in the partial sticky-price case. Verbally describe
what is going on.

---

## Solution

### Background: the shock and the long run

$\theta_t$ is the **disutility-of-labor** parameter. A permanent rise in
$\theta_t$ shifts the labor supply curve $N^s(w_t, \theta_t)$ permanently to
the left: workers are less willing to supply any given number of hours at any
given real wage. Because $\theta_t$ does not enter the IS curve, the LM curve,
or the production function, its only direct effect is on labor supply.

**Long-run (Neoclassical) outcome** — the equilibrium that both models converge
to once prices have fully adjusted:

| Variable | Direction | Reason |
|----------|-----------|--------|
| $Y_t^{ncl}$ | $\downarrow$ | lower labor supply → lower employment → lower output |
| $P_t^{ncl}$ | $\uparrow$ | lower real money demand (lower $Y$, higher $r$) at fixed $M$ |
| $r_t^{ncl}$ | $\uparrow$ | lower $Y^s$ moves up the IS curve |
| $w_t^{ncl}$ | $\uparrow$ | labor-demand / shifted-labor-supply intersection |
| $N_t^{ncl}$ | $\downarrow$ | labor-supply contraction |

The question is how quickly — and along what path — each model reaches this
long-run.

---

### Part (a) — Simple Sticky-Price Model

**Mechanism.** The AS curve is the horizontal line $P_t = \bar P_{0,t}$,
exogenous and independent of $\theta_t$. The AD curve is also unaffected
($\theta_t$ enters neither IS nor LM). Therefore:

- **Periods 1–3 (sticky-price regime):** The AS–AD intersection is *unchanged*.
  Output stays at $Y_{0,t}$, the price level stays at $\bar P_{0,t}$, and
  because both $Y_t$ and $P_t$ are unchanged the IS–LM equilibrium is also
  unchanged: $r_t = r_{0,t}$.
- The only variable that adjusts is the **real wage**: with employment pinned at
  $N_{0,t}$ (output unchanged, production function unchanged) but labor supply
  shifted left, the wage must rise to $w_{1,t}$ to induce the same amount of
  labor at the higher disutility.
- **Period 3+ (Neoclassical long run):** Prices fully adjust. Output falls to
  $Y_t^{ncl}$, the price level rises to $P_t^{ncl}$, and the real interest
  rate rises to $r_t^{ncl}$.

**Key insight.** Nominal rigidities *completely* insulate output, prices, and
the interest rate from a supply shock in the short and medium run. All
adjustment in periods 1–3 is borne by the real wage alone.

```{r}
#| label: fig-ex271-ssp
#| fig-cap: "Simple sticky-price model: permanent $\\theta_t$ increase. Horizontal lines are solid; vertical event markers are dotted. $\\theta_t$ rises at period 1 and stays elevated. Output, prices, and the real interest rate are unchanged through the sticky-price regime (periods 1–3); all three jump to their Neoclassical long-run levels from period 3 onward."
#| fig-alt: "Four time-series panels for the SSP theta shock. Theta steps up at period 1 and stays high. Y, P, and r are flat during periods 1-3 then jump to NCL levels at period 3."
#| fig-height: 5.5
#| fig-width: 8.5

# θ: permanent jump at t = 1; stays at θ₁ forever
seg_th_a <- make_hsegs(c(theta0, theta1), breaks = 1)

# Y: flat at Y₀ through t = 3, then falls to Y_ncl
seg_Y_a  <- make_hsegs(c(Y0, Y0, Y_ncl),  breaks = c(1, 3))

# P: flat at P₀ through t = 3, then rises to P_ncl
seg_P_a  <- make_hsegs(c(P0, P0, P_ncl),  breaks = c(1, 3))

# r: flat at r₀ through t = 3, then rises to r_ncl
seg_r_a  <- make_hsegs(c(r0, r0, r_ncl),  breaks = c(1, 3))

p_th_a <- plot_tp(
  seg_th_a,
  ylab       = expression(theta[t]),
  ref_labels = list(
    list(y = theta0, lab = expression(theta["0,t"])),
    list(y = theta1, lab = expression(theta["1,t"]))
  ),
  vert_lines = c(1, 3)
)

p_Y_a <- plot_tp(
  seg_Y_a,
  ylab       = expression(Y[t]),
  ref_labels = list(
    list(y = Y_ncl, lab = expression(Y[t]^ncl)),
    list(y = Y0,    lab = expression(Y["0,t"]))
  ),
  vert_lines = c(1, 3)
)

p_P_a <- plot_tp(
  seg_P_a,
  ylab       = expression(P[t]),
  ref_labels = list(
    list(y = P0,    lab = expression(bar(P)["0,t"])),
    list(y = P_ncl, lab = expression(P[t]^ncl))
  ),
  vert_lines = c(1, 3)
)

p_r_a <- plot_tp(
  seg_r_a,
  ylab       = expression(r[t]),
  ref_labels = list(
    list(y = r0,    lab = expression(r["0,t"])),
    list(y = r_ncl, lab = expression(r[t]^ncl))
  ),
  vert_lines = c(1, 3)
)

grid.arrange(p_th_a, p_Y_a,
             p_P_a,  p_r_a,
             nrow = 2, ncol = 2)
```

---

### Part (b) — Partial Sticky-Price Model

In the partial sticky-price model prices adjust gradually. The AS curve is

$$P_t = \bar{P}_t + \gamma\!\left(Y_t - Y_t^f\right), \qquad \gamma > 0,$$

where $Y_t^f$ is the Neoclassical (flex-price) output level. A higher $\theta_t$
lowers $Y_t^f$, which shifts the AS curve **left** (upward).

**Mechanism.**

1. $\theta_t \uparrow$ permanently $\Rightarrow$ $Y_t^f \downarrow$
   $\Rightarrow$ AS shifts left. AD unchanged.
2. **Periods 1–2 (partial adjustment, halfway to NCL):** the new AS–AD
   intersection lies halfway between the original equilibrium and the
   Neoclassical long-run values. Output falls to $Y_t^{mr}$, prices rise to
   $P_t^{mr}$, and the higher price level contracts real balances, pushing
   the real interest rate up to $r_t^{mr}$.
3. **Period 3+ (full NCL adjustment):** prices complete their adjustment.
   The economy settles at $Y_t^{ncl}$, $P_t^{ncl}$, $r_t^{ncl}$ — the same
   long-run destination as in the simple sticky-price model.

**Labor market.** At every step, lower output requires less labor ($N_t
\downarrow$). The real wage rises at each stage because workers are on the new,
leftward $N^s$ curve: $w_t \uparrow$ throughout the adjustment.

**Stagflation.** The combination of falling output and rising prices in periods
1–2 is the hallmark of a supply-shock stagflation: the leftward AS shift moves
the economy up and to the left along the unchanged AD curve.

**Comparison with part (a).** In the SSP model output and prices are completely
insulated from the supply shock in periods 1–3; adjustment is deferred entirely
to the long run. In the PSP model there is partial but immediate pass-through:
output falls and prices rise in periods 1–2, reaching the Neoclassical
equilibrium one period earlier (from period 3). Nominal rigidities
*dampen and delay* the real output losses from supply shocks but cannot
prevent them entirely.

```{r}
#| label: fig-ex271-psp
#| fig-cap: "Partial sticky-price model: permanent $\\theta_t$ increase. Horizontal lines are solid; vertical event markers are dotted. $\\theta_t$ rises at period 1 and stays elevated. Output, prices, and the real interest rate adjust halfway toward the Neoclassical long-run in periods 1–2 ($Y_t^{mr}$, $P_t^{mr}$, $r_t^{mr}$), then complete the adjustment from period 3 onward ($Y_t^{ncl}$, $P_t^{ncl}$, $r_t^{ncl}$)."
#| fig-alt: "Four time-series panels for the PSP theta shock. Theta steps up permanently at period 1. Y falls in two steps: halfway at period 1, full NCL at period 3. P and r rise in two steps in the same fashion."
#| fig-height: 5.5
#| fig-width: 8.5

# θ: same permanent jump as part (a)
seg_th_b <- make_hsegs(c(theta0, theta1), breaks = 1)

# Y: drops halfway at t=1 (periods 1–2), drops to NCL at t=3
seg_Y_b  <- make_hsegs(c(Y0, Y_mr, Y_ncl),  breaks = c(1, 3))

# P: rises halfway at t=1, rises to NCL at t=3
seg_P_b  <- make_hsegs(c(P0, P_mr, P_ncl),  breaks = c(1, 3))

# r: rises halfway at t=1, rises to NCL at t=3
seg_r_b  <- make_hsegs(c(r0, r_mr, r_ncl),  breaks = c(1, 3))

p_th_b <- plot_tp(
  seg_th_b,
  ylab       = expression(theta[t]),
  ref_labels = list(
    list(y = theta0, lab = expression(theta["0,t"])),
    list(y = theta1, lab = expression(theta["1,t"]))
  ),
  vert_lines = c(1, 3)
)

p_Y_b <- plot_tp(
  seg_Y_b,
  ylab       = expression(Y[t]),
  ref_labels = list(
    list(y = Y_ncl, lab = expression(Y[t]^ncl)),
    list(y = Y_mr,  lab = expression(Y[t]^mr)),
    list(y = Y0,    lab = expression(Y["0,t"]))
  ),
  vert_lines = c(1, 3)
)

p_P_b <- plot_tp(
  seg_P_b,
  ylab       = expression(P[t]),
  ref_labels = list(
    list(y = P0,    lab = expression(P["0,t"])),
    list(y = P_mr,  lab = expression(P[t]^mr)),
    list(y = P_ncl, lab = expression(P[t]^ncl))
  ),
  vert_lines = c(1, 3)
)

p_r_b <- plot_tp(
  seg_r_b,
  ylab       = expression(r[t]),
  ref_labels = list(
    list(y = r0,    lab = expression(r["0,t"])),
    list(y = r_mr,  lab = expression(r[t]^mr)),
    list(y = r_ncl, lab = expression(r[t]^ncl))
  ),
  vert_lines = c(1, 3)
)

grid.arrange(p_th_b, p_Y_b,
             p_P_b,  p_r_b,
             nrow = 2, ncol = 2)
```

**Summary — cross-model comparison.**

| Variable | SSP periods 1–3 | PSP periods 1–2 | Both period 3+ |
|----------|:---------------:|:---------------:|:--------------:|
| $Y_t$ | unchanged | $\downarrow$ (halfway) | $\downarrow$ to $Y_t^{ncl}$ |
| $P_t$ | unchanged | $\uparrow$ (halfway) | $\uparrow$ to $P_t^{ncl}$ |
| $r_t$ | unchanged | $\uparrow$ (halfway) | $\uparrow$ to $r_t^{ncl}$ |
| $w_t$ | $\uparrow$ (large) | $\uparrow$ (moderate) | $\uparrow$ to $w_t^{ncl}$ |
| $N_t$ | unchanged | $\downarrow$ (halfway) | $\downarrow$ to $N_t^{ncl}$ |
