---
title: "ECON 503 Textbook Exercises — Chapter 5"
author: "Pedro Amaral"
date: today
format:
  html:
    toc: true
    code-fold: show
    css: style_C.css
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)

# Custom ggplot theme matching the document CSS (Option C: warm off-white + navy)
theme_econ503 <- theme_minimal(base_size = 14, base_family = "serif") +
  theme(
    plot.background   = element_rect(fill = "#fdfcf8", color = NA),
    panel.background  = element_rect(fill = "#fdfcf8", color = NA),
    panel.grid.major  = element_line(color = "#e2ddd4", linewidth = 0.3),
    panel.grid.minor  = element_line(color = "#eeeae0", linewidth = 0.2),
    plot.title        = element_text(color = "#1a3a6b", face = "bold", size = 16),
    axis.title        = element_text(color = "#2c5282"),
    axis.text         = element_text(color = "#4a5568"),
    legend.background = element_rect(fill = "#fdfcf8", color = NA),
    legend.text       = element_text(color = "#4a5568"),
    legend.position   = "bottom"
  )

# Set as default theme for all plots
theme_set(theme_econ503)
```

---

# Exercise 5.1 — CES Production Function

**Problem.** Suppose that the production function is:
$$Y_t = A\left[\alpha K_t^{\frac{\nu-1}{\nu}} + (1-\alpha)N_t^{\frac{\nu-1}{\nu}}\right]^{\frac{\nu}{\nu-1}}$$
with $\nu \geq 0$ and $0 < \alpha < 1$.

**(a)** Prove that this production function features constant returns to scale.

**(b)** Compute the first partial derivatives with respect to $K_t$ and $N_t$. Argue that these are positive.

**(c)** Compute the own second partial derivatives with respect to $K_t$ and $N_t$. Show that these are both negative.

**(d)** As $\nu \to 1$, how do the first and second partial derivatives for this production function compare with the Cobb-Douglas production function discussed in the text?

---

## Solution

**Notation.** Define $Z_t \equiv \alpha K_t^{\frac{\nu-1}{\nu}} + (1-\alpha)N_t^{\frac{\nu-1}{\nu}}$ so that $Y_t = AZ_t^{\frac{\nu}{\nu-1}}$.

### Part (a) — Constant returns to scale

A production function has **constant returns to scale** if $F(\lambda K_t, \lambda N_t) = \lambda F(K_t, N_t)$ for all $\lambda > 0$. Substituting:

$$F(\lambda K_t, \lambda N_t) = A\!\left[\alpha(\lambda K_t)^{\frac{\nu-1}{\nu}} + (1-\alpha)(\lambda N_t)^{\frac{\nu-1}{\nu}}\right]^{\frac{\nu}{\nu-1}}$$

$$= A\!\left[\lambda^{\frac{\nu-1}{\nu}}\!\left(\alpha K_t^{\frac{\nu-1}{\nu}} + (1-\alpha)N_t^{\frac{\nu-1}{\nu}}\right)\right]^{\frac{\nu}{\nu-1}}$$

$$= A\,\lambda^{\frac{\nu-1}{\nu}\cdot\frac{\nu}{\nu-1}} Z_t^{\frac{\nu}{\nu-1}} = \lambda \cdot AZ_t^{\frac{\nu}{\nu-1}} = \lambda\, Y_t. \quad \checkmark$$

The exponent on $\lambda$ simplifies to exactly 1, confirming CRS. Note that **diminishing marginal products are not contradicted by CRS**: CRS describes how output scales when *both* inputs are doubled together, while diminishing marginal products describe how output changes when *one* input increases holding the other fixed. These are independent properties.

### Part (b) — First partial derivatives

Using the chain rule with $Y_t = AZ_t^{\frac{\nu}{\nu-1}}$:

$$F_K \equiv \frac{\partial Y_t}{\partial K_t} = A\cdot\frac{\nu}{\nu-1}\,Z_t^{\frac{\nu}{\nu-1}-1}\cdot\alpha\cdot\frac{\nu-1}{\nu}\,K_t^{\frac{\nu-1}{\nu}-1} = A\alpha\, Z_t^{\frac{1}{\nu-1}}\,K_t^{-\frac{1}{\nu}}$$

$$F_N \equiv \frac{\partial Y_t}{\partial N_t} = A(1-\alpha)\, Z_t^{\frac{1}{\nu-1}}\,N_t^{-\frac{1}{\nu}}$$

**Both are positive** since $A > 0$, $\alpha \in (0,1)$, and all of $Z_t^{1/(\nu-1)}$, $K_t^{-1/\nu}$, $N_t^{-1/\nu}$ are strictly positive for $K_t, N_t > 0$ (a positive number raised to any real power remains positive).

### Part (c) — Own second partial derivatives

Differentiating $F_K = A\alpha Z_t^{\frac{1}{\nu-1}} K_t^{-\frac{1}{\nu}}$ again with respect to $K_t$:

$$F_{KK} = A\alpha\!\left[\frac{1}{\nu-1}Z_t^{\frac{1}{\nu-1}-1}\cdot\alpha\frac{\nu-1}{\nu}K_t^{-\frac{1}{\nu}}\cdot K_t^{-\frac{1}{\nu}} + Z_t^{\frac{1}{\nu-1}}\cdot\left(-\frac{1}{\nu}\right)K_t^{-\frac{1}{\nu}-1}\right]$$

$$= \frac{A\alpha}{\nu}\!\left[\alpha\, Z_t^{\frac{2-\nu}{\nu-1}}\,K_t^{-\frac{2}{\nu}} - Z_t^{\frac{1}{\nu-1}}\,K_t^{-\frac{1}{\nu}-1}\right]$$

Factoring out $\frac{A\alpha}{\nu} Z_t^{\frac{2-\nu}{\nu-1}} K_t^{-\frac{2}{\nu}}$ (which is **positive**):

$$F_{KK} = \frac{A\alpha}{\nu}\,Z_t^{\frac{2-\nu}{\nu-1}}\,K_t^{-\frac{2}{\nu}}\!\left[\alpha - Z_t\,K_t^{\frac{1}{\nu}-1}\right]$$

To determine the sign of the bracket, expand $Z_t K_t^{\frac{1}{\nu}-1}$:

$$Z_t\,K_t^{\frac{1}{\nu}-1} = \alpha\,K_t^{\frac{\nu-1}{\nu}+\frac{1}{\nu}-1} + (1-\alpha)\,N_t^{\frac{\nu-1}{\nu}}\,K_t^{\frac{1}{\nu}-1} = \alpha + (1-\alpha)\!\left(\frac{N_t}{K_t}\right)^{\!\frac{\nu-1}{\nu}} > \alpha$$

where the last equality uses $\frac{\nu-1}{\nu}+\frac{1}{\nu}-1 = 0$ and the last inequality holds because $(1-\alpha)\left(N_t/K_t\right)^{(\nu-1)/\nu}>0$.

Therefore $\alpha - Z_t K_t^{\frac{1}{\nu}-1} < 0$, which gives $F_{KK} < 0$. By symmetry (swapping the roles of $K_t$, $\alpha$ with $N_t$, $1-\alpha$), $F_{NN} < 0$. $\checkmark$

### Part (d) — Limit as $\nu \to 1$ (Cobb-Douglas)

**Cobb-Douglas derivatives for reference.** For $F^{CD} = AK_t^{\alpha}N_t^{1-\alpha}$:
$$F_K^{CD} = \alpha AK_t^{\alpha-1}N_t^{1-\alpha}, \qquad F_{KK}^{CD} = \alpha(\alpha-1)AK_t^{\alpha-2}N_t^{1-\alpha}$$
and by symmetry (swap $K_t,\alpha \leftrightarrow N_t,1-\alpha$):
$$F_N^{CD} = (1-\alpha)AK_t^{\alpha}N_t^{-\alpha}, \qquad F_{NN}^{CD} = -\alpha(1-\alpha)AK_t^{\alpha}N_t^{-\alpha-1}$$

**Useful identity.** Since $Y_t = AZ_t^{\nu/(\nu-1)}$, raising both sides to the power $1/\nu$ gives $Z_t^{1/(\nu-1)} = (Y_t/A)^{1/\nu}$. This lets us eliminate $Z_t$ from the derivative expressions.

**Limit of $F_K$.** Substituting the identity into the expression from Part (b):
$$F_K = A\alpha\,Z_t^{\frac{1}{\nu-1}}\,K_t^{-\frac{1}{\nu}} = A\alpha\left(\frac{Y_t}{A}\right)^{\!\frac{1}{\nu}}K_t^{-\frac{1}{\nu}} = \alpha\,A^{1-\frac{1}{\nu}}\!\left(\frac{Y_t}{K_t}\right)^{\!\frac{1}{\nu}}$$

As $\nu\to 1$: $A^{1-1/\nu}\to 1$ and $(Y_t/K_t)^{1/\nu}\to Y_t/K_t$, so:
$$\lim_{\nu\to 1}F_K = \frac{\alpha\,Y_t}{K_t} \;\stackrel{*}{=}\; \alpha AK_t^{\alpha-1}N_t^{1-\alpha} = F_K^{CD} \checkmark$$

**Limit of $F_{KK}$.** Write $(2-\nu)/(\nu-1) = 1/(\nu-1)-1$, so $Z_t^{(2-\nu)/(\nu-1)} = (Y_t/A)^{1/\nu}/Z_t$. As $\nu\to 1$: $Z_t\to 1$, $K_t^{-2/\nu}\to K_t^{-2}$, $Z_t^{(2-\nu)/(\nu-1)}\to Y_t/A$, and the bracket from Part (c):
$$\alpha - Z_t K_t^{\frac{1}{\nu}-1} \;\xrightarrow{\;\nu\to 1\;}\; \alpha - 1\cdot K_t^{0} = -(1-\alpha)$$
Therefore, from the expression in Part (c):
$$\lim_{\nu\to 1}F_{KK} = A\alpha\cdot\frac{Y_t}{A}\cdot K_t^{-2}\cdot[-(1-\alpha)] = -\alpha(1-\alpha)\frac{Y_t}{K_t^2} \;\stackrel{*}{=}\; \alpha(\alpha-1)AK_t^{\alpha-2}N_t^{1-\alpha} = F_{KK}^{CD} \checkmark$$

By symmetry the same steps give $\lim_{\nu\to 1}F_N = F_N^{CD}$ and $\lim_{\nu\to 1}F_{NN} = F_{NN}^{CD}$.

$$\lim_{\nu\to 1}F_K = F_K^{CD}, \qquad \lim_{\nu\to 1}F_{KK} = F_{KK}^{CD}$$

*(Steps marked $\stackrel{*}{=}$ use $\lim_{\nu\to1}Y_t = AK_t^\alpha N_t^{1-\alpha}$, proved directly below.)*

**Limit of the production function itself.** As $\nu \to 1$, the inner exponent $\frac{\nu-1}{\nu} \to 0$, so $K_t^{(\nu-1)/\nu} \to 1$ and $N_t^{(\nu-1)/\nu} \to 1$; the bracket converges to 1 while the outer exponent $\frac{\nu}{\nu-1} \to \infty$ — a $1^\infty$ indeterminate form.

**Taking logarithms** converts it to a $\tfrac{0}{0}$ form. Starting from $Y_t = A[\cdots]^{\nu/(\nu-1)}$:
$$\ln Y_t = \ln A + \frac{\ln\!\left[\alpha K_t^{\frac{\nu-1}{\nu}} + (1-\alpha)N_t^{\frac{\nu-1}{\nu}}\right]}{\dfrac{\nu-1}{\nu}}$$

**Apply L'Hôpital's rule** (numerator $\to\ln 1=0$, denominator $\to 0$). The denominator's derivative is $1/\nu^2$; the numerator's derivative is:
$$\frac{1}{\nu^2}\cdot\frac{\alpha K_t^{\frac{\nu-1}{\nu}}\ln K_t + (1-\alpha)N_t^{\frac{\nu-1}{\nu}}\ln N_t}{\alpha K_t^{\frac{\nu-1}{\nu}} + (1-\alpha)N_t^{\frac{\nu-1}{\nu}}}$$

The $1/\nu^2$ cancels. Evaluating at $\nu=1$ (so all $X_t^{(\nu-1)/\nu}\to 1$):
$$\lim_{\nu\to 1}\ln Y_t = \ln A + \frac{\alpha\ln K_t + (1-\alpha)\ln N_t}{\alpha + (1-\alpha)} = \ln A + \alpha\ln K_t + (1-\alpha)\ln N_t$$

Exponentiating:
$$\lim_{\nu \to 1} Y_t = AK_t^{\alpha}\,N_t^{1-\alpha}$$

The CES production function nests the **Cobb-Douglas** $AK_t^\alpha N_t^{1-\alpha}$ as the special case $\nu = 1$. Since the production function and all its first and second partial derivatives converge to their Cobb-Douglas counterparts, the two specifications are locally identical at $\nu = 1$.

---

# Exercise 5.2 — Labor Force Doubling in the Solow Model

**Problem.** Suppose that you have a standard Solow model. The central equation governing the dynamics of the *level* of capital is:
$$K_{t+1} = sAF(K_t, N_t) + (1-\delta)K_t$$
In terms of capital *per worker*, the central equation is:
$$k_{t+1} = sAf(k_t) + (1-\delta)k_t$$
The production function has the normal properties.

**(a)** Suppose that the economy initially sits in a steady state in terms of the capital stock per worker, $k_t$. Suppose that, at time $t$, the number of workers doubles (due to an influx of immigrants). The number of workers is expected to remain forever constant at this new higher level, i.e., $N_{t+1} = N_t$. Graphically analyze how this will impact the steady state capital stock per worker and the dynamics starting from an initial capital stock.

**(b)** Draw diagrams plotting out how capital, output, and the real wage ought to respond dynamically to the permanent increase in the workforce.

---

## Solution

### Part (a) — Graphical analysis

**The steady state is unchanged in per-worker terms.** The central equation $k_{t+1} = sAf(k_t) + (1-\delta)k_t$ depends only on per-worker capital, the saving rate, TFP, and the depreciation rate — not on the level of $N$. Since $s$, $A$, and $\delta$ are all unchanged, the steady state $k^*$ satisfying $\delta k^* = sAf(k^*)$ is the same before and after the immigration shock.

**What changes is the initial condition.** At time $t$, the capital stock $K_t = k^* N$ is predetermined (capital cannot jump). When $N$ doubles to $2N$, capital per worker immediately halves:
$$k_t = \frac{K_t}{2N} = \frac{k^*\! N}{2N} = \frac{k^*}{2}$$

The economy is now below its (unchanged) steady state and converges back to $k^*$ from below, exactly as if it had started with insufficient capital.

```{r ex2_solow_diagram, fig.cap="Solow diagram: doubling of the workforce drops $k_t$ to $k^*/2$; the economy then converges back to the same $k^*$.", fig.alt="Solow diagram: k_{t+1} vs k_t. Curve crosses 45° line at k*; red line at k*/2 marks post-shock capital."}
alpha <- 1/3; s <- 0.20; A <- 1; delta <- 0.10

k_star <- (s * A / delta)^(1 / (1 - alpha))
k_seq  <- seq(0.001, 2 * k_star, length.out = 400)

df_solow <- tibble(
  k      = k_seq,
  k_next = s * A * k_seq^alpha + (1 - delta) * k_seq,
  invest = s * A * k_seq^alpha,
  deprec = delta * k_seq
)

ggplot(df_solow, aes(x = k)) +
  geom_line(aes(y = k_next), color = "steelblue", linewidth = 1, linetype = "solid") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = k_star,     linetype = "dotted", color = "darkgreen") +
  geom_vline(xintercept = k_star / 2, linetype = "dotted", color = "firebrick") +
  annotate("text", x = k_star + 0.05,     y = 0.1, label = "k*",    color = "darkgreen", hjust = 0) +
  annotate("text", x = k_star / 2 + 0.05, y = 0.1, label = "k*/2",  color = "firebrick", hjust = 0) +
  annotate("text", x = 1.55 * k_star, y = 1.55 * k_star - 0.08,
           label = "45° line", color = "gray40", size = 3) +
  labs(x = expression(k[t]), y = expression(k[t+1]),
       title = "Central equation of the Solow model",
       subtitle = "Immigration shock drops capital per worker to k*/2; convergence back to k*")
```

### Part (b) — Dynamic paths

```{r ex2_dynamics, fig.cap="Dynamic responses to a permanent doubling of the workforce at period 10.", fig.height=6, fig.alt="Time series: k, y, w per worker (normalized to 1). All drop sharply at period 10, then recover to 1."}
T   <- 60
t0  <- 10   # period of shock

# Simulate per-worker capital path
k <- numeric(T)
k[1:(t0 - 1)] <- k_star          # in steady state before shock
k[t0]         <- k_star / 2      # shock: N doubles, K predetermined

for (t in t0:(T - 1)) {
  k[t + 1] <- s * A * k[t]^alpha + (1 - delta) * k[t]
}

# Derived series
y <- A * k^alpha                            # output per worker
w <- (1 - alpha) * A * k^alpha              # real wage (= (1-alpha)*y for C-D)

df_dyn <- tibble(
  period = 1:T,
  `Capital per worker`  = k / k_star,       # normalized to pre-shock s.s. = 1
  `Output per worker`   = y / (A * k_star^alpha),
  `Real wage`           = w / ((1 - alpha) * A * k_star^alpha)
) |>
  pivot_longer(-period, names_to = "variable", values_to = "value")

ggplot(df_dyn, aes(x = period, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = t0, linetype = "dashed", color = "gray50") +
  annotate("text", x = t0 + 0.5, y = 1.02, label = "Shock", color = "gray40",
           hjust = 0, size = 3.5) +
  scale_color_manual(values = c("steelblue", "darkorange", "darkgreen")) +
  labs(x = "Period", y = "Value (normalized to pre-shock steady state = 1)",
       title = "Dynamic response to a permanent doubling of the workforce",
       color = NULL)
```

**Interpretation.** At period $t_0$ the number of workers doubles instantaneously:

- **Capital per worker** $k_t$ drops to $k^*/2$ on impact (capital is predetermined), then gradually rises back to $k^*$.
- **Output per worker** $y_t = Af(k_t)$ mirrors $k_t$: it drops on impact and recovers over time.
- **Real wage** $w_t = Af(k_t) - Af'(k_t)k_t$ also drops on impact. With Cobb-Douglas, $w_t = (1-\alpha)y_t$, so it moves proportionally with output per worker and recovers to its original steady-state level.

Note that aggregate output $Y_t = y_t \cdot N_t$ jumps *up* on impact (even though per-worker output falls) because labor force doubles. In the new steady state, aggregate capital and output are both double their original level, while all per-worker variables return to their original values.

---

# Exercise 5.3 — Solow Model with a Government

**Problem.** Suppose that we have a Solow model with one twist: there is a government. Each period, the government consumes a fraction of output, $s_G$. Hence the aggregate resource constraint is:
$$Y_t = C_t + I_t + G_t$$
where $G_t = s_G Y_t$. Define **private output** as $Y_t^p = Y_t - G_t$. Suppose that investment is a constant fraction $s$ of private output (consumption is then $1-s$ times private output). Otherwise the model is the same as in the text.

**(a)** Re-derive the central equation of the Solow model under this setup.

**(b)** Suppose that the economy initially sits in a steady state. Suppose that there is an increase in $s_G$ that is expected to last forever. Graphically analyze how this will affect the steady state value of the capital stock per worker. Plot out a graph showing how the capital stock per worker will be affected in a dynamic sense.

---

## Solution

### Part (a) — Re-deriving the central equation

Since $G_t = s_G Y_t$, private output is $Y_t^p = (1-s_G)Y_t$. Investment is a fixed fraction of private output:
$$I_t = s Y_t^p = s(1-s_G)Y_t = s(1-s_G) A F(K_t, N_t)$$

The capital accumulation equation is unchanged:
$$K_{t+1} = I_t + (1-\delta)K_t = s(1-s_G)AF(K_t, N_t) + (1-\delta)K_t$$

Dividing through by $N$ (constant population) and using $y_t = Af(k_t)$:

$$k_{t+1} = s(1-s_G)\,Af(k_t) + (1-\delta)k_t$$

This is identical in form to the standard Solow central equation but with an **effective saving rate** $\tilde{s} \equiv s(1-s_G)$. Government spending reduces the resources available for private investment and therefore acts like a lower saving rate.

The steady state satisfies $\delta k^* = s(1-s_G)Af(k^*)$. For the Cobb-Douglas case $f(k)=k^\alpha$:
$$k^* = \left(\frac{s(1-s_G)A}{\delta}\right)^{\!\frac{1}{1-\alpha}}$$

**Note on private consumption:** $c_t = (1-s)Y_t^p = (1-s)(1-s_G)Af(k_t)$, so steady-state consumption per worker is:
$$c^* = (1-s)(1-s_G)\,Af(k^*)$$

### Part (b) — Graphical analysis and dynamics

An increase in $s_G$ reduces the effective saving rate $\tilde{s} = s(1-s_G)$, which **shifts down** the gross investment schedule $\tilde{s}Af(k_t)$. This results in a lower steady-state capital stock per worker $k_1^* < k_0^*$.

At time $t_0$ when $s_G$ rises, capital $k_{t_0}$ is predetermined and cannot jump. Since investment is immediately lower, depreciation exceeds investment and $k$ falls over time toward the new, lower steady state $k_1^*$.

```{r ex3_dynamics, fig.cap="Solow diagram and dynamic path after a permanent increase in the government spending share.", fig.width=10, fig.height=5, fig.alt="Left: Solow diagram with lower investment after sG rise. Right: k per worker declining to new lower steady state."}
alpha <- 1/3; A <- 1; delta <- 0.10; s <- 0.20
sG0 <- 0.10; sG1 <- 0.25   # government spending share before and after

k_star0 <- ((s * (1 - sG0) * A) / delta)^(1 / (1 - alpha))
k_star1 <- ((s * (1 - sG1) * A) / delta)^(1 / (1 - alpha))

# --- Panel 1: Solow diagram ---
k_max <- 1.2 * k_star0
k_seq <- seq(0.001, k_max, length.out = 400)

# scale_color_manual assigns alphabetically: "After"->steelblue, "Before"->firebrick, "Depreciation"->black
p1 <- tibble(k = k_seq,
       invest0 = s * (1 - sG0) * A * k_seq^alpha,
       invest1 = s * (1 - sG1) * A * k_seq^alpha,
       deprec  = delta * k_seq) |>
  ggplot(aes(x = k)) +
  geom_line(aes(y = invest0, color = "Before (low sG)"), linewidth = 1) +
  geom_line(aes(y = invest1, color = "After (high sG)"),  linewidth = 1) +
  geom_line(aes(y = deprec,  color = "Depreciation"),     linewidth = 1) +
  geom_vline(xintercept = k_star0, linetype = "dotted", color = "firebrick") +
  geom_vline(xintercept = k_star1, linetype = "dotted", color = "steelblue") +
  annotate("text", x = k_star0 + 0.04, y = 0.003,
           label = 'k[0]^"*"', parse = TRUE, color = "firebrick", hjust = 0) +
  annotate("text", x = k_star1 - 0.04, y = 0.003,
           label = 'k[1]^"*"', parse = TRUE, color = "steelblue", hjust = 1) +
  scale_color_manual(values = c("steelblue","firebrick","black")) +
  labs(x = expression(k[t]), y = "Investment / Depreciation",
       title = "Increase in sG  lowers investment",
       color = NULL) +
  theme(legend.position = "bottom")
# --- Panel 2: Dynamic path ---
T  <- 60; t0 <- 10
k  <- numeric(T)
k[1:t0] <- k_star0   # initialize through t0 so the loop starts from the old steady state
for (t in t0:(T - 1)) {
  k[t + 1] <- s * (1 - sG1) * A * k[t]^alpha + (1 - delta) * k[t]
}

p2 <- tibble(period = 1:T, k = k) |>
  ggplot(aes(x = period, y = k)) +
  geom_line(color = "gray30", linewidth = 1) +
  geom_hline(yintercept = k_star0, linetype = "dashed", color = "firebrick") +
  geom_hline(yintercept = k_star1, linetype = "dashed", color = "steelblue") +
  geom_vline(xintercept = t0, linetype = "dotted", color = "gray40") +
  annotate("text", x = 55, y = k_star0 + 0.02,
           label = 'k[0]^"*"', parse = TRUE, color = "firebrick") +
  annotate("text", x = 55, y = k_star1 - 0.02,
           label = 'k[1]^"*"', parse = TRUE, color = "steelblue") +
  labs(x = "Period", y = expression(k[t]),
       title = "Dynamic path of capital per worker",
       subtitle = "Permanent increase in sG at period 10")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

# Exercise 5.4 — Golden Rule in the Cobb-Douglas Solow Model

**Problem.** Suppose that we have a standard Solow model with a Cobb-Douglas production function. The central equation of the model is:
$$k_{t+1} = sAk_t^{\alpha} + (1-\delta)k_t$$
Consumption per worker is given by:
$$c_t = (1-s)Ak_t^{\alpha}$$

**(a)** Solve for an expression for the steady state capital stock per worker. Assume that productivity is fixed at $A$.

**(b)** Use your answer in part (a) to solve for an expression for steady state consumption per worker.

**(c)** Use calculus to derive an expression for the $s$ which maximizes steady state consumption per worker.

---

## Solution

### Part (a) — Steady state capital per worker

At steady state $k_{t+1} = k_t = k^*$:
$$k^* = sAk^{*\alpha} + (1-\delta)k^* \implies \delta k^* = sAk^{*\alpha} \implies k^{*1-\alpha} = \frac{sA}{\delta}$$

$$k^* = \left(\frac{sA}{\delta}\right)^{\!\frac{1}{1-\alpha}}$$

As expected, $k^*$ is increasing in the saving rate $s$ and in TFP $A$, and decreasing in the depreciation rate $\delta$.

### Part (b) — Steady state consumption per worker

Substituting $k^*$ into $c^* = (1-s)Ak^{*\alpha}$:

$$k^{*\alpha} = \left(\frac{sA}{\delta}\right)^{\!\frac{\alpha}{1-\alpha}}$$

$$c^* = (1-s)\,A\!\left(\frac{sA}{\delta}\right)^{\!\frac{\alpha}{1-\alpha}}$$

### Part (c) — Golden Rule saving rate

We maximize $c^*$ with respect to $s$. Taking the derivative and setting it to zero:

$$\frac{dc^*}{ds} = -A\!\left(\frac{sA}{\delta}\right)^{\!\frac{\alpha}{1-\alpha}} + (1-s)\,A\cdot\frac{\alpha}{1-\alpha}\!\left(\frac{sA}{\delta}\right)^{\!\frac{\alpha}{1-\alpha}-1}\!\cdot\frac{A}{\delta} = 0$$

Factoring out $A\!\left(\frac{sA}{\delta}\right)^{\!\frac{\alpha}{1-\alpha}}$:

$$-1 + (1-s)\,\frac{\alpha}{(1-\alpha)\,s} = 0 \implies \frac{\alpha(1-s)}{s(1-\alpha)} = 1$$

Solving for $s$:

$$\alpha - \alpha s = s(1-\alpha) \implies \alpha = s\bigl[\alpha + 1 - \alpha\bigr] = s$$

$$s^{GR} = \alpha$$

The **Golden Rule saving rate** equals the capital income share $\alpha$.

**Verification via the Golden Rule condition.** The textbook characterizes the Golden Rule by $Af'(k^*) = \delta$, i.e., the marginal product of capital equals the depreciation rate. With Cobb-Douglas:
$$Af'(k^*) = \alpha A k^{*\alpha-1} = \alpha A\left(\frac{sA}{\delta}\right)^{-1} = \frac{\alpha\delta}{s}$$

Setting this equal to $\delta$ gives $s = \alpha$ $\checkmark$

```{r ex4_golden_rule, fig.cap="Steady-state consumption per worker as a function of the saving rate. Maximum is at $s = \\alpha$.", fig.alt="Hump-shaped c* vs s curve, peaking at s = alpha = 1/3 (red dashed line). Higher or lower s reduces steady-state consumption."}
alpha <- 1/3; A <- 1; delta <- 0.10

s_seq <- seq(0.001, 0.999, length.out = 500)
k_ss  <- (s_seq * A / delta)^(1 / (1 - alpha))
c_ss  <- (1 - s_seq) * A * k_ss^alpha

tibble(s = s_seq, c_star = c_ss) |>
  ggplot(aes(x = s, y = c_star)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = alpha, linetype = "dashed", color = "firebrick") +
  annotate("text", x = alpha + 0.02, y = 0.02,
           label = expression(s^GR == alpha), color = "firebrick", hjust = 0) +
  labs(x = "Saving rate s",
       y = expression(paste("Steady-state consumption per worker", " ", c^"*")),
       title = "Golden Rule: steady-state consumption is maximized at s = α",
       subtitle = paste0("Parameters: α = ", round(alpha, 2), ", A = ", A, ", δ = ", delta))
```

**Intuition.** At $s = \alpha$, the marginal product of capital exactly covers depreciation. If $s < \alpha$, a small increase in saving raises steady-state output more than it raises steady-state investment, raising $c^*$. If $s > \alpha$, the economy is dynamically inefficient (saving too much): reducing $s$ unambiguously raises $c^*$ in both the short and long run.
